---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# stdmatern

This directory is for development of fast and memory-efficient code that creates Matérn precision matrices that have been standardized so that their inverse is a correlation matrix. The code is written in C++ and made available inside R with the `{Rcpp}` packages.

The package can be installed with 

```{r}
#| eval: false
pak::pak("bgautijonsson/stdmatern")
```



```{r}
library(stdmatern)
Q <- make_standardized_matern(dim = 2, rho = 0.5, nu = 0)
```

```{r}
Q
```

```{r}
Q |> solve()
```

Creating and standardizing a 1600x1600 precision matrix

```{r}
bench::mark(
  make_standardized_matern(dim = 40, rho = 0.5, nu = 0)
)
```


# Sampling spatial data


```{r}
library(dplyr)
library(sparseMVN)
library(ggplot2)

Q <- make_standardized_matern(dim = 20, rho = 0.9, nu = 2)

grid_dim <- 50
rho <- 0.9
nu <- 2
Q <- make_standardized_matern(grid_dim, rho, nu = nu)


Z <- rmvn.sparse(
    n = 1,
    mu = rep(0, nrow(Q)),
    CH = Matrix::Cholesky(Q)
)

tibble(
  Z = as.numeric(Z)
) |> 
  mutate(
    id = row_number(),
      lat = (id - 1) %% grid_dim,
      lon = cumsum(lat == 0),
    ) |> 
    ggplot(aes(lat, lon, fill = Z)) +
    geom_raster() +
    scale_fill_viridis_c() +
    coord_fixed(expand = FALSE)
```